---
title: "Interactive widgets: a tour"
description: >
  Everything gloss adds on top of a static plot: hover tooltips (including HTML),
  click-select and brush, pan/zoom by mouse/touch/keyboard, the toolbar and
  export, linked views, and styling.
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = requireNamespace("quill", quietly = TRUE)
)
library(gloss)
```

`gloss` turns a `vellum` scene — or, more usually, a `quill` plot — into a
self-contained, client-side interactive HTML widget. There is one verb,
`as_widget()`, and it works the same in the RStudio/Positron viewer, a
knitr/Quarto document, and a Shiny app. No Shiny server and no round-trip are
required; every interaction runs in the browser.

Interactivity is *declared in the grammar*: a mark's `data_id`, `tooltip`, and
`hover_group` arguments flow through the scene as per-element metadata that the
widget reads. A plot that declares none still renders as a static (but
embeddable) SVG.

```{r}
#| label: setup
library(quill)
df <- data.frame(
  wt = mtcars$wt, mpg = mtcars$mpg,
  model = rownames(mtcars), cyl = factor(mtcars$cyl)
)
```

## Hover: tooltips and highlighting

Map `tooltip` and `data_id` and hover a point: it shows a tooltip and dims the
others. `data_id` is the join key; `tooltip` is what the box says.

```{r}
#| label: hover
vplot(df) |>
  mark_point(x = wt, y = mpg, color = cyl, tooltip = model, data_id = model) |>
  as_widget()
```

### HTML tooltips

Tooltip text is rendered as **safe HTML**: build a multi-line, formatted string
in `tooltip =` (here with `glue`) using `<b>` and `<br>`. Data values are
escaped, and only inert tags (`<b>`, `<i>`, `<br>`, `<span>`) are honoured, so
there is no injection risk. Style the box with `tooltip_style`.

```{r}
#| label: html-tooltip
df$label <- glue::glue("<b>{df$model}</b><br>{df$mpg} mpg · {df$wt}k lbs")
vplot(df) |>
  mark_point(x = wt, y = mpg, tooltip = label, data_id = model) |>
  as_widget(tooltip_style = list(background = "#1d3557", fontsize = "13px"))
```

## Select and brush

Click a mark to select it (every mark sharing its `data_id` toggles together);
drag a rectangle to brush-select. `select_mode = "single"` makes a click replace
the selection instead of toggling.

```{r}
#| label: brush
vplot(df) |>
  mark_point(x = wt, y = mpg, data_id = model, selected_color = "#e63946") |>
  as_widget()
```

## Pan, zoom, and the toolbar

Zoom with the mouse wheel, drag to pan (toggle brush/pan on the toolbar). On a
touch device, drag pans and a two-finger pinch zooms; with the widget focused,
the arrow keys pan, `+`/`-` zoom, and `0` resets. The on-hover toolbar adds
zoom-to-selection, reset, SVG/PNG download, copy-to-clipboard (where supported),
and fullscreen.

Export captures the *current* view, so a zoomed-in region exports as shown. Set
the download name and a hi-res PNG scale:

```{r}
#| label: export
vplot(df) |>
  mark_point(x = wt, y = mpg, data_id = model) |>
  as_widget(export_filename = "mtcars", export_scale = 2)
```

## Linked views

Widgets sharing a `group` link client-side: selecting or brushing in one
highlights the same data keys in the others — no Shiny, no crosstalk. Selection
projects by `hover_group` when the marks declare it.

```{r}
#| label: linked
p <- vplot(df) |> mark_point(x = wt, y = mpg, data_id = model, hover_group = cyl)
htmltools::tagList(
  as_widget(p, group = "cars", width = 320, height = 240),
  as_widget(vplot(df) |> mark_point(x = wt, y = mpg, data_id = model, hover_group = cyl),
            group = "cars", width = 320, height = 240)
)
```

For interop with plotly / leaflet / DT and crosstalk's `filter_*` controls, pass
a `crosstalk::SharedData` (or a group name) to `crosstalk =` instead.

## Styling

`hover_color`, `selected_color`, and `dim_opacity` set the widget-wide theme;
per-mark `hover_color` / `selected_color` declared in `quill` override them for
that mark. `tooltip_style` themes the tooltip box (above).

```{r}
#| label: styling
vplot(df) |>
  mark_point(x = wt, y = mpg, data_id = model) |>
  as_widget(hover_color = "seagreen", selected_color = "firebrick", dim_opacity = 0.15)
```

## Shiny

`glossOutput()` / `renderGloss()` embed a widget in a Shiny app (display-tier
today; server-side selection read-back is planned). See `?gloss-shiny`.
