---
title: "Using vellumwidget in Shiny"
description: >
  Embed a vellumwidget widget in a Shiny app and react to what the user does — read the
  selection, clicks, hovers, and brush as reactive inputs.
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(vellumwidget)
```

A `vellumwidget` widget is interactive on its own (hover, select, brush, pan/zoom) with
no server round-trip. Inside Shiny it also **reports those interactions back** as
reactive inputs, so an app can respond to what the user does on the plot: filter a
table to the brushed points, show details for the hovered mark, drive another
output from the selection.

## The two bindings

Render a widget with the standard htmlwidgets pair, exactly like `plotly` or `DT`:

```{r}
#| eval: false
library(shiny)
library(vellumplot)
library(vellumwidget)

ui <- fluidPage(
  vellumwidgetOutput("plot", height = "500px"),
  verbatimTextOutput("info")
)

server <- function(input, output) {
  output$plot <- renderVellumwidget({
    vplot(mtcars) |>
      mark_point(
        x = wt, y = mpg,
        tooltip = rownames(mtcars),
        data_id = rownames(mtcars) # <- the key each interaction reports
      ) |>
      as_widget()
  })

  output$info <- renderPrint({
    input$plot_selected # the keys currently selected
  })
}

shinyApp(ui, server)
```

## What the widget reports

The inputs are named `<outputId>_<event>`, so for `vellumwidgetOutput("plot")` they are
`input$plot_selected`, `input$plot_click`, `input$plot_hover`,
`input$plot_brush`, and `input$plot_zoom`. **Every selection value is expressed
in element keys** (the `data_id` you set on the mark), so you map back to your
data by that key.

| Input | Value | Kind |
|---|---|---|
| `input$plot_selected` | character vector of selected keys (`character(0)` if none) | state |
| `input$plot_click` | `list(key=)` — `key` is `NULL` for an empty-space click | event |
| `input$plot_hover` | the hovered key, or `NULL` on leave | state |
| `input$plot_brush` | `list(keys=, x0=, y0=, x1=, y1=)` — keys + region rectangle (device px); a lasso adds `lasso=TRUE`; a cartesian plot adds data-space bounds `x0d,y0d,x1d,y1d` + `panel` | event |
| `input$plot_zoom` | `list(x=, y=, w=, h=, zoomed=)` — the current view (viewBox, device px) + a zoomed flag; a single-panel cartesian plot adds `data=list(x=, y=, panel=)` (visible range in data coordinates) | state |

Data-space fields (`x0d`/… and `zoom$data`) appear only when the plot carries a
cartesian scale (any `vellumplot` plot); a raw `vellum` scene reports device-pixel
fields only. They describe the **visual** axes — `x0d`/`x1d` is the horizontal
axis, which under `coord_flip()` is the plot's `y` aesthetic. Date/time axes report
the numeric epoch (days for `Date`, seconds for `POSIXct`) — map back with
`as.Date()` / `.POSIXct()`. A discrete axis reports fractional band positions, and
an axis with a custom `scales::transform_*()` (beyond identity / log10 / sqrt /
reverse) is omitted from the data-space fields.

**State vs event.** *State* inputs (`_selected`, `_hover`, `_zoom`) update only
when the value changes, so re-selecting the same set (or settling on the same
view) does not re-fire; pair them with `observe()` / `reactive()`. *Event* inputs
(`_click`, `_brush`) fire on every occurrence, even a repeat; pair them with
`observeEvent()`.

## Reacting to interactions

Because everything is keyed, the common patterns are one-liners. Filter a table to
the brushed points:

```{r}
#| eval: false
output$table <- renderTable({
  keys <- input$plot_brush$keys
  if (length(keys)) mtcars[rownames(mtcars) %in% keys, ] else mtcars[0, ]
})
```

Show the hovered car:

```{r}
#| eval: false
output$detail <- renderText({
  k <- input$plot_hover
  if (is.null(k)) "Hover a point" else paste(k, "-", mtcars[k, "hp"], "hp")
})
```

Respond to each click:

```{r}
#| eval: false
observeEvent(input$plot_click, {
  key <- input$plot_click$key
  if (!is.null(key)) showNotification(paste("clicked", key))
})
```

## Driving the widget from the server

Reading interactions is one direction; the other is the server *changing* what
the widget shows. `vellumwidget_proxy()` gives you a handle to a widget that is
**already on the page** and drives it **without re-rendering** — the SVG is not
rebuilt, so the current pan/zoom and the smooth feel are preserved. It works
exactly like `leaflet::leafletProxy()`, `DT::dataTableProxy()`, or
`plotly::plotlyProxy()`.

Call it with the same `outputId` you gave `vellumwidgetOutput()`, then pipe the
handle through the verbs:

| Verb | Effect |
|---|---|
| `vw_select(keys)` / `vw_clear_selection()` | set / clear the selection |
| `vw_filter(keys)` / `vw_clear_filter()` | cross-filter to `keys` (dim the rest) / remove the filter |
| `vw_zoom(keys)` / `vw_reset_zoom()` | frame `keys` / restore the full view |

All `keys` are the element `data_id`s — the very same identifiers you get back
through `input$plot_selected`. For example, a `selectInput` that highlights cars
on the plot, and a button that clears everything:

```{r}
#| eval: false
ui <- fluidPage(
  selectInput("pick", "Highlight", choices = rownames(mtcars), multiple = TRUE),
  actionButton("reset", "Reset"),
  vellumwidgetOutput("plot", height = "500px")
)

server <- function(input, output, session) {
  output$plot <- renderVellumwidget({
    vplot(mtcars) |>
      mark_point(x = wt, y = mpg, data_id = rownames(mtcars)) |>
      as_widget()
  })

  # server -> client: highlight and zoom to the picked cars, no re-render
  observeEvent(input$pick, ignoreNULL = FALSE, {
    vellumwidget_proxy("plot") |>
      vw_select(input$pick) |>
      vw_zoom(input$pick) # empty selection resets the view
  })

  observeEvent(input$reset, {
    vellumwidget_proxy("plot") |>
      vw_clear_selection() |>
      vw_reset_zoom()
  })
}
```

A server-driven `vw_select()` behaves just like a user click: it projects across
a mark's `hover_group`, propagates to any linked / crosstalk widgets, and updates
`input$plot_selected`. If you also *observe* `input$plot_selected` and call
`vw_select()` from there, guard against the echo (e.g. only act when the value
actually differs) to avoid a feedback loop — the usual Shiny proxy caution.

## Notes

- The `data_id` (mark key) is what everything reports; without it the marks are
  drawn but carry no identity to report. `tooltip` is independent; it is what the
  hover box shows, not what the input carries.
- The inputs fire only inside a live Shiny session. A static render (knitr,
  pkgdown, `htmltools::save_html()`) is byte-for-byte identical and emits nothing.
- The same keys are what a [crosstalk][ct] `SharedData` uses, so a vellumwidget widget can
  link to `DT`, `plotly`, or `leaflet` client-side *without* Shiny. See
  `as_widget(crosstalk=)`. Shiny inputs and crosstalk are complementary: use Shiny
  when the server needs to react, crosstalk for pure client-side linking.
- Driving the widget *from* the server — selection, filter, zoom — is
  `vellumwidget_proxy()`, above. Shiny inputs (client → server), crosstalk (client ↔
  client), and the proxy (server → client) are the three linking directions.

[ct]: https://rstudio.github.io/crosstalk/

See the [interactive widgets tour](interactivity.html) for the full set of
client-side interactions, and the
[accessibility article](https://r-vellum.github.io/vellumplot/articles/accessibility.html)
for the keyboard and screen-reader model.
