---
title: "Using vellumwidget in Shiny"
description: >
  Embed a vellumwidget widget in a Shiny app and react to what the user does — read the
  selection, clicks, hovers, and brush as reactive inputs.
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(vellumwidget)
```

A `vellumwidget` widget is interactive on its own — hover, select, brush, pan/zoom — with
no server round-trip. Inside Shiny it also **reports those interactions back** as
reactive inputs, so an app can respond to what the user does on the plot: filter a
table to the brushed points, show details for the hovered mark, drive another
output from the selection.

## The two bindings

Render a widget with the standard htmlwidgets pair, exactly like `plotly` or `DT`:

```{r}
#| eval: false
library(shiny)
library(vellumplot)
library(vellumwidget)

ui <- fluidPage(
  vellumwidgetOutput("plot", height = "500px"),
  verbatimTextOutput("info")
)

server <- function(input, output) {
  output$plot <- renderVellumwidget({
    vplot(mtcars) |>
      mark_point(
        x = wt, y = mpg,
        tooltip = rownames(mtcars),
        data_id = rownames(mtcars) # <- the key each interaction reports
      ) |>
      as_widget()
  })

  output$info <- renderPrint({
    input$plot_selected # the keys currently selected
  })
}

shinyApp(ui, server)
```

## What the widget reports

The inputs are named `<outputId>_<event>` — so for `vellumwidgetOutput("plot")` they are
`input$plot_selected`, `input$plot_click`, `input$plot_hover`, and
`input$plot_brush`. **Every value is expressed in element keys** — the `data_id`
you set on the mark — so you map back to your data by that key.

| Input | Value | Kind |
|---|---|---|
| `input$plot_selected` | character vector of selected keys (`character(0)` if none) | state |
| `input$plot_click` | `list(key=)` — `key` is `NULL` for an empty-space click | event |
| `input$plot_hover` | the hovered key, or `NULL` on leave | state |
| `input$plot_brush` | `list(keys=, x0=, y0=, x1=, y1=)` — keys + brushed rectangle (device px) | event |

**State vs event.** *State* inputs (`_selected`, `_hover`) update only when the
value changes, so re-selecting the same set does not re-fire — pair them with
`observe()` / `reactive()`. *Event* inputs (`_click`, `_brush`) fire on every
occurrence, even a repeat — pair them with `observeEvent()`.

## Reacting to interactions

Because everything is keyed, the common patterns are one-liners. Filter a table to
the brushed points:

```{r}
#| eval: false
output$table <- renderTable({
  keys <- input$plot_brush$keys
  if (length(keys)) mtcars[rownames(mtcars) %in% keys, ] else mtcars[0, ]
})
```

Show the hovered car:

```{r}
#| eval: false
output$detail <- renderText({
  k <- input$plot_hover
  if (is.null(k)) "Hover a point" else paste(k, "-", mtcars[k, "hp"], "hp")
})
```

Respond to each click:

```{r}
#| eval: false
observeEvent(input$plot_click, {
  key <- input$plot_click$key
  if (!is.null(key)) showNotification(paste("clicked", key))
})
```

## Notes

- The `data_id` (mark key) is what everything reports; without it the marks are
  drawn but carry no identity to report. `tooltip` is independent — it is what the
  hover box shows, not what the input carries.
- The inputs fire only inside a live Shiny session. A static render (knitr,
  pkgdown, `htmltools::save_html()`) is byte-for-byte identical and emits nothing.
- The same keys are what a [crosstalk][ct] `SharedData` uses, so a vellumwidget widget can
  link to `DT`, `plotly`, or `leaflet` client-side *without* Shiny — see
  `as_widget(crosstalk=)`. Shiny inputs and crosstalk are complementary: use Shiny
  when the server needs to react, crosstalk for pure client-side linking.
- Driving the widget *from* the server — setting the selection or a filter without
  re-rendering — is a planned addition (a `vellumwidget_proxy()`), not yet available.

[ct]: https://rstudio.github.io/crosstalk/

See the [interactive widgets tour](interactivity.html) for the full set of
client-side interactions, and the
[accessibility article](https://r-vellum.github.io/vellumplot/articles/accessibility.html)
for the keyboard and screen-reader model.
